Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex
cleanup

\noindent {\bf Under-approximating the environment for non-termination.}
Dual to termination, when proving non-termination, we need to under-approximate the environment and the loop's body and apply
formula {\bf [CNT]}. %As opposed to over-approximations 
%this is not as straightforward as when computing over-approximations in the termination case described above.
Recall that there is one more constraint that we must obey when doing so: we must stay inside the synthesis fragment! 
This becomes more challenging now then when dealing with over-approximations due to the existential nature of under-approximations: they require an additional existential quantifier, which might result in an additional quantifier alternation, i.e. $\exists \forall \exists$.

This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level.

However, when dealing with nested loops, we need to  under-approximate the inner loop, requiring a nested existential quantifier, 
resulting in the $\exists \forall \exists$ alternation.  At first sight, this seems fatal for our formulation.  Luckily, it is possible to
reformulate non-termination as an $\exists \forall$ formula by observing that, unlike a ranking function, the defining property of a
recurrence set is \emph{non relational} -- if we end up in the recurrence set, we don't care exactly where we came from as long as
we know that it was also somewhere in the recurrence set.  This allows us to cast non-termination of nested loops as the formula
shown in Figure~\ref{fig:nonterm-nested}.

If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
\begin{itemize}
\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
\item If $L_2$ is non-terminating, then $N_2 \wedge G_2$ is its recurrence set.
\end{itemize}

>>>>

\noindent {\bf Under-approximating the environment for non-termination.}
Dual to termination, when proving non-termination, we need to
under-approximate the environment and the loop's body and apply
formula {\bf [CNT]}. %As opposed to over-approximations 
%this is not as straightforward as when computing over-approximations in the termination case described above.
Recall that there is one more constraint that we must obey when doing so: we
must stay inside the synthesis fragment!  This becomes more challenging now
then when dealing with over-approximations due to the existential nature of
under-approximations: they require an additional existential quantifier,
which might result in an additional quantifier alternation, i.e.~$\exists
\forall \exists$.

This issue does not arise when abstracting the environment, as its
under-approximation and the loop under consideration make use of existential
quantification at the same nestedness level.

However, when dealing with nested loops, we need to under-approximate the
inner loop, requiring a nested existential quantifier, resulting in the
$\exists \forall \exists$ alternation.  At first sight, this seems fatal for
our formulation.  Luckily, it is possible to reformulate non-termination as
an $\exists \forall$ formula by observing that, unlike a ranking function,
the defining property of a recurrence set is \emph{non relational} -- if we
end up in the recurrence set, we do not care exactly where we came from as
long as we know that it was also somewhere in the recurrence set.  This
allows us to cast non-termination of nested loops as the formula shown in
Figure~\ref{fig:nonterm-nested}.

If the formula on the right-hand side of the figure is satisfiable, then
$L_1$ is non-terminating, as witnessed by the recurrence set $N_1$ and the
initial state $x_0$ in which the program begins executing.  There are two
possible scenarios for $L_2$'s termination:
%
\begin{itemize}
%
\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that
reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x)
\wedge P_2(x,x') \rightarrow N_1(x') $.
%
\item If $L_2$ is non-terminating, then $N_2 \wedge G_2$ is its recurrence set.
%
\end{itemize}
